---
fontsize: 12pt
geometry: margin=0.25in
header-includes:
    - \usepackage{caption}
title: 'WAS cell lineage study - supplemental 1'
author: 'John K. Everett'
output:
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
---
  
```{r, echo=FALSE, eval=FALSE}
library(tidyverse)
options(stringsAsFactors = FALSE)

# Read in intSite data. The intSite positions and abundances are the result of processing the output 
# of the INSPIIRED intSite pipeline with C Berry's dRep software package.
intSites <- read.csv(gzfile('../data/intSites.csv.gz'), header = TRUE)
for(i in c('start', 'end', 'width', 'estAbund')) intSites[[i]] <- as.numeric(intSites[[i]]) 
intSites[which(is.na(intSites$nearestFeature)),]$nearestFeature <- 'NONE'


# Read in the list of samples that are to be exported to the SRA.
SRA_samples <- read.table('../data/SRA_samples.txt', sep = ';', header = TRUE)

# For plot clarity, change all m12_6 time points to m12. This only pertains to paient FR04.
intSites$timepoint <- as.character(intSites$timepoint)
intSites[which(intSites$timepoint == 'm12_6'),]$timepoint <- 'm12'

# Restrict data set to SRA samples.
all(SRA_samples$SpecimenAccNum %in% intSites$sampleName)
intSites <- subset(intSites, sampleName %in% SRA_samples$SpecimenAccNum)

# Convert the timepoint character string to a numeric days column.
intSites$timePointDays <- toupper(intSites$timepoint)
intSites$timePointDays <- gsub('_', '.', intSites$timePointDays)
intSites$timePointDays <- sapply(intSites$timePointDays, function(x){
  if(grepl('D', x)){
    return(as.numeric(stringr::str_match(x, '[\\d\\.]+')))
  } else if(grepl('M', x)){
    return(as.numeric(stringr::str_match(x, '[\\d\\.]+'))*30.4167)
  } else if(grepl('Y', x)){
    return(as.numeric(stringr::str_match(x, '[\\d\\.]+'))*365)
  } else {
    message('Warning - could not determine date unit for: ', x)
    return(as.numeric(stringr::str_match(x, '[\\d\\.]+')))
  }
})

# Rename cell types.
intSites$celltype <- as.character(intSites$celltype)
intSites$celltype <- gsub('NEUTROPHILS', 'GRANULOCYTES', intSites$celltype)  

# Change the patient identifiers.
convertName <- function(x){
  nameConversion <- data.frame(internal=c('FR01', 'FR03', 'FR04', 'FR05', 'LT',    'p1204', 'TST', 'p1202', 'All'), 
                               external=c('WAS2', 'WAS4', 'WAS5', 'WAS7', 'bS/bS', 'bS/bS', 'b0/bE', 'b0/bE', 'All')) 
  
  nameConversion[match(tolower(x), tolower(nameConversion$internal)),]$external
}

intSites$patient <- convertName(intSites$patient)

# Consider patient position ids since the same position may occasionally be found between subjects.
intSites$patientPosid <- paste(intSites$patient, intSites$posid)

d <- data.frame(intSites)
save.image(file = 'Everett_oncoGeneEnrichment_1.RData')
```

```{r, echo=FALSE, message=FALSE, warnings=FALSE, error=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(gt23)
options(stringsAsFactors = FALSE)
load('Everett_oncoGeneEnrichment_1.RData')

ppNum   <- function(n) format(n, big.mark = ",", scientific = FALSE, trim = TRUE)
percent <- function(x, digits = 2, format = "f", ...) paste0(formatC(100 * x, format = format, digits = digits, ...), "%")
  
testOncoGenes <- function(distCol){
  o <- list()
  o[['m']] <- matrix(c(n_distinct(d[d$timePointDays <   690 & abs(d[,distCol]) >   50000,]$patientPosid),
                       n_distinct(d[d$timePointDays <   690 & abs(d[,distCol]) <=  50000,]$patientPosid),
                       n_distinct(d[d$timePointDays >=  690 & abs(d[,distCol]) >   50000,]$patientPosid),
                       n_distinct(d[d$timePointDays >=  690 & abs(d[,distCol]) <=  50000,]$patientPosid)), 
                       byrow = TRUE, nrow=2)  

  o[['f']] <- fisher.test(o[['m']])
  dimnames(o[['m']]) <- list(c('First time point', 'Second time point'), c('nSites > 50 KB onocogenes', 'nSites $\\leq$ 50 KB onocogenes'))
  o[['early']] <- percent(o[['m']][1,2] / sum(o[['m']][1,]))
  o[['late']]  <- percent(o[['m']][2,2] / sum(o[['m']][2,]))
 
  sets <- split(d, d$patient)
  o[['r']] <- bind_rows(mapply(function(x, n){
    m <- matrix(c(n_distinct(x[x$timePointDays <   690 & abs(x[,distCol]) >   50000,]$patientPosid),
                  n_distinct(x[x$timePointDays <   690 & abs(x[,distCol]) <=  50000,]$patientPosid),
                  n_distinct(x[x$timePointDays >=  690 & abs(x[,distCol]) >   50000,]$patientPosid),
                  n_distinct(x[x$timePointDays >=  690 & abs(x[,distCol]) <=  50000,]$patientPosid)), 
                  byrow = TRUE, nrow=2)  

    data.frame(subject=n,
               earlyPercentNearOnco = (m[1,2]/sum(m[1,]))*100,
               earlySites   =  paste0(ppNum(m[1,2]), ' / ', ppNum(m[1,1])),
               latePercentNearOnco  = (m[2,2]/sum(m[2,]))*100,
               lateSites   = paste0(ppNum(m[2,2]), ' / ', ppNum(m[2,1])))
  }, sets, names(sets), SIMPLIFY=FALSE))


  o[['w']] <- wilcox.test(o[['r']]$earlyPercentNearOnco, o[['r']]$latePercentNearOnco, paired = TRUE)
  
  o[['r']]$earlyPercentNearOnco <- paste0(round(o[['r']]$earlyPercentNearOnco, 2), '%')
  o[['r']]$latePercentNearOnco <- paste0(round(o[['r']]$latePercentNearOnco, 2), '%')
  names(o[['r']]) <- c('Subject', 'TP1 percent sites near oncogenes', 'TP1 sites near / far\nfrom oncogenes', 'TP2 percent sites near oncogenes', 'TP2 sites near / far\nfrom oncogenes')
  o
}

fullOncoListTests <- testOncoGenes('nearestOncoFeatureDist')
lymphomaOncoListTests <- testOncoGenes('nearestlymphomaFeatureDist')
```
\captionsetup[table]{labelformat=empty}

**Full list of oncogenes considered (`r ppNum(n_distinct(gt23::humanOncoGenesList))` genes)**  
When considering all subjects, a Fisher's exact test of unique intSites near oncogenes, comparing first time points 
(`r fullOncoListTests$early` intSites near oncogenes) and second time points (`r fullOncoListTests$late` intSites near oncogenes), 
showed a small yet significant depletion of intSites near oncogenes in the second time points 
(p-value: `r formatC(fullOncoListTests$f$p.value, format = 'f', digits = 4)`).  

```{r, echo=FALSE}
m <- ppNum(fullOncoListTests$m)
kable(m) 
```

For each of the six subjects, comparing the percentages of intSites near oncogenes from the first and second time points showed no significant difference 
(paired Wilcoxcon test, p-value: `r formatC(fullOncoListTests$w$p.value, format = 'f', digits = 3)`).  

```{r, echo=FALSE}
row.names(fullOncoListTests$r) <- NULL
kable(fullOncoListTests$r, format = "latex", align = 'c', booktabs = T, linesep = "") %>%
  column_spec(2:5, width = "4cm") %>%
  kable_styling(latex_options = c("striped"))
```

\vspace{0.5cm}

**Lymphoma associated oncogenes considered (`r ppNum(n_distinct(gt23::humanLymphomaGenesList))` genes)**  
When considering all subjects, a Fisher's exact test of unique intSites near oncogenes, comparing first time points 
(`r lymphomaOncoListTests$early` intSites near oncogenes) and second time points (`r lymphomaOncoListTests$late` intSites near oncogenes), 
no significant difference was found (p-value: `r formatC(lymphomaOncoListTests$f$p.value, format = 'f', digits = 2)`).  
  
```{r, echo=FALSE}
m <- ppNum(lymphomaOncoListTests$m)
kable(m)
```

For each of the six subjects, comparing the percentages of intSites near oncogenes from the first and second time points showed no significant difference 
(paired Wilcoxcon test, p-value: `r formatC(lymphomaOncoListTests$w$p.value, format = 'f', digits = 3)`).

```{r, echo=FALSE}
row.names(lymphomaOncoListTests$r) <- NULL
kable(lymphomaOncoListTests$r, format = "latex", align = 'c', booktabs = T, linesep = "") %>%
  column_spec(2:5, width = "4cm") %>%
  kable_styling(latex_options = c("striped"))
```

\newpage

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(RMySQL)
invisible(sapply(DBI::dbListConnections(MySQL()), dbDisconnect))
dbConn <- dbConnect(MySQL(), group = 'intSites_miseq')

o <- bind_rows(lapply(SRA_samples$SpecimenAccNum, function(x){
  run <- unname(unlist(dbGetQuery(dbConn,  paste0("select distinct miseqid from samples where sampleName like '", x, "-%'"))))
  uninfectedReps <- length(unname(unlist(dbGetQuery(dbConn,  paste0("select distinct sampleName from samples where sampleName like 'UninfectedControl%' and miseqid='", run, "'")))))
  notemplateReps <- length(unname(unlist(dbGetQuery(dbConn,  paste0("select distinct sampleName from samples where sampleName like 'NoTemplateControl%' and miseqid='", run, "'")))))
  uninfectedReps <- ifelse(length(uninfectedReps) == 0, 0, uninfectedReps)
  notemplateReps <- ifelse(length(notemplateReps) == 0, 0, notemplateReps)
  
  uninfected  <- unname(unlist(dbGetQuery(dbConn,  paste0("select distinct siteId from sites where sampleID in (select sampleID from samples where sampleName like 'UninfectedControl%' and miseqid='", run, "')"))))
  notemplate <- unname(unlist(dbGetQuery(dbConn,  paste0("select distinct siteId from sites where sampleID in (select sampleID from samples where sampleName like 'NoTemplateControl%' and miseqid='", run, "')"))))
  uninfected <- ifelse(length(uninfected) == 0, 0, length(uninfected))
  notemplate <- ifelse(length(notemplate) == 0, 0, length(notemplate))
  
  uninfected <- ifelse(uninfectedReps == 0, NA, uninfected)
  notemplate <- ifelse(notemplateReps == 0, NA, notemplate)
  
  data.frame(Sample = x, 
             'Run' = run, 
             'Unifected_control_replicates' = uninfectedReps, 
             'Unifected_control_sites' = uninfected, 
             'No_DNA_control_replicates' = notemplateReps,
             'No_DNA_control_sites' = notemplate, 
             check.names = FALSE)
}))

row.names(o) <- NULL

o2 <- group_by(o, Run) %>%
      summarise('Number of samples' = n_distinct(Sample),
                'Uninfected control' = ifelse(sum(Unifected_control_replicates) > 0, 'Yes', 'No'),
                'Uninfected control sites' = ifelse(sum(Unifected_control_replicates) > 0, sum(Unifected_control_sites), NA),
                'No DNA control' = ifelse(sum(No_DNA_control_replicates) > 0, 'Yes', 'No'),
                'No DNA control sites' = ifelse(sum(No_DNA_control_replicates) > 0, sum(No_DNA_control_sites), NA))
```

The table below shows the controls and results of those controls for each sequencing run used in this study. Four technical replicates 
where employed for each control experiment. Two intSites were found in the uninfected control experiment for sequencing run 160426_M03249_0159_000000000-AM2PA 
which included a single study sample (WAS7, m12/GRANULOCYTES, 879 intSites).  

\vspace{1.0cm}

```{r, echo=FALSE}
o2 %>% kable(format = "latex", align='lccccc', booktabs = T, linesep = "", caption = "Sequencing experiment controls") %>%
  column_spec(1, width = "7cm") %>%
  column_spec(2:6, width = "2cm") %>%
kable_styling(font_size = 10, latex_options = c("hold_position", "striped"))
```

